\bf Comparing the Power of Plot Designs to Reveal Correlation

### 2016-01-08 paper/paper.Rnw
Cleanup

<<setup, echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
library(lme4)
library(ggplot2)
library(mnormt)
library(tidyr)
library(knitr)
# Convert reshape2 to tidyr, melt to gather - done
#library(reshape2)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)
library(stringr)
library(ipapi)
library(maps)
library(moments)
#library(pander)
#library(splitstackshape) - changed cSplit to unnest
#library(readr)
#library(lubridate)
# Convert these to read_csv at some point
correl <- read.csv("correlations.csv", stringsAsFactors = FALSE)
full <- read.csv("full.csv", row.names=1, stringsAsFactors = FALSE) 
Turk <- read.csv("turk.csv", stringsAsFactors = FALSE)

@@

To decide on an appropriate display requires awareness of cognitive principles in psychology such as the Gestalt law of common fate described by \citet{Wertheimer}. This law describes a tendency in human perception to 'see' a positive relationship between two objects that briefly move together over time, leading people to believe that they stay linked and share a 'common fate' (Figure~\ref{cny}), even if no relationship exists. \citet{Lee} work finds that humans are able to perceive a relationship between objects that move with temporal synchrony in accordance with the Gestalt Law, implying that a graphic than can utilise a dimension for time may be able to use the information provided by temporal dependence and potentially outperform the graph types that do not, such as the scatter plot. However, keep in mind magical thinking \citep{Diaconis},   the desire to find any pattern may lead to a false discovery of positive correlation. The law then has two effects, if positive correlation exists in temporal data, the observer may be drawn to it easier. If positive correlation does not exist, the observer may be fooled into thinking it does anyway. Either effect would lead to a more confident rejection of the null than if the same data was displayed as a scatter plot. Figure~\ref{cny} highlights the dangerous interaction between 'magical thinking' and the Gestalt Law, the observer is drawn to the period from April 8 to April 22, and may falsely believe that the two currencies may continue to move together. However, over the entire series the linear correlation between the two currencies is -0.12, indicating there is no actual relationship. Incorrectly relying on two currencies, or many other sets of economic variables, being dependent on each other can be extremely dangerous for a firm. \citet{Robbins} is similarly critical of the time-based line graph, where it is argued that information from overlaid line graphs can easily be misleading (Figure~\ref{murder}). When both lines dive after 1993, it is natural to compare the horizontal distance and say that the difference between the lines in 1994 is small. However, this is incorrect, as looking at the small horizontal distance is comparing the lines at different dates. It is not obvious that the difference in 1994 is the second greatest difference across the entire range, only slightly smaller than in 1993. \citet[Vanderplas] examines this effect in more detail. It is dubbed the 'Sine Illusion', where the human eye is poor at interpreting lines with such a sharp slope. As these problems are unique to the line graph, they argue that the scatter plot, which does not have any ability to display time, is the superior choice. However, in practice many data analysts are split between the two major alternatives for temporal data. The overlaid line graph is justified by its ability to present more information to the analyst; but many argue that it is this extra information that is misleading, and revert to the scatter plot for its strong non-temporal performance.

@@

\begin{equation} \label{eq:1}
  \centering
  p-value = P(X \geq x | H_{0}) = 1 - B_{n, 1/m}(x-1)
\end{equation}

If all $K$ observers pick the same plot out of a lineup of twenty, we result in a p-value as small as $1/m^K$ . This research used $m= 20$, giving us a significance level (and Type 1 error rate) of $α = 0.05$. The plot of 'true' data is placed randomly amongst 19 null plots to form the lineup.

If a subject selects multiple plots from the lineup, the p-value has a Bernoulli distribution with a probability equal to $s/m$, where that observer made $s$ selections out of a lineup with $m$ plots \citep{Follett}. In the event that a lineup has observers making different numbers of selections, we have a binomial distribution with different success probabilities, and a Poisson Binomial distribution must be used to calculate the p-value. The probability mass function of a Poisson Binomial distribution can be written as:

$$P(X=x) = \sum_{A \in T_{x}} \prod_{j \in A} p_{j} \prod_{j \in A^{c}} (1-p_{j})$$

@@

As the power of a statistical test is defined as the probability of rejecting $H_{0}$ when it is false, the power of a lineup test is viewed as the probability of detecting the true plot. We approximate the power as $\hat\pi = x/K$, where $x$ observers out of $K$ correctly detected the true plot out of the lineup.
We can them estimate the power difference of competing lineups, $\hat\pi_{1} - \hat\pi_{2}$. An $\alpha \times 100\%$ confidence interval is calculated as \citep{Hoffman:2012}:

\begin{equation} \label{eq:2}
  \centering
  \hat\pi_{1} - \hat\pi_{2}\pm t_{1-\alpha,2n-1}\sqrt{\hat\pi_{1} (1-\hat\pi_{1}) /n_{1} + \hat\pi_{2} (1-\hat\pi_{2}) /n_{2} }
\end{equation}

@@

Some individuals may have a better natural ability at detecting the correct correlation pattern in the graphics, but by having each participant viewing multiple lineups and each lineup being viewed by multiple different participants, this can be controlled via a random effect variable in the model. We recruited participants through Amazon's Mechanical Turk, where the lineup protocol has been applied to a number of problems in prior papers \citep{Heer, Majumder:2013, Wickham}. Amazon’s Mechanical Turk \citep{MTurk} is a labour crowd-sourcing platform developed to give easy access to workers with basic tasks paid in line with the United States minimum wage. MTurk can be used to recruit subjects to read a variety of graphics and report on particular visual tasks. The time taken to decide, confidence in the decision and the reason for that decision can also be recorded. \citet{Heer} use MTurk to replicate previous findings in graphical perception from \citet{Cleveland} and find that it is a valid method of data collection, once controls to eliminate anyone 'gaming' the system by randomly selecting answers to minimise time spent working are implemented. Further details of our use of Amazons’ Mechanical Turk is included in Section 3. The protocol has successfully been used to measure statistical power as a means to determine plot type superiority in \citet{Hoffman:2012}, and this work follows that approach.

@@

$$~N(\mu= \left| \begin{array}{c}0 \\0 \end{array} \right|,\hspace{4mm}\Sigma = \left| \begin{array}{cc}1 & \rho \\ \rho & 1\end{array} \right|)$$

The variation in the covariance parameter  allows for control in the correlation in the pairs of data. The null data has no structure, so uncorrelated pairs are produced by setting $\rho = 0$. True data pairs are generated with $\rho = \pm 0.3, 0.5, 0.7, 0.9$. Pilot testing found that the rate of picking the true data plot fell to approximately $5\%$, the $1/m$ rate of randomly picking the true plot, for pairs with correlation below $\rho = 0.3$. Further to this, relationships weaker than this point are often of little interest to analysts. Hence we found it unnecessary to test a more full range of correlations. Simulating data does not produce correlations exactly equal to $\rho$, so imulated data pairs were accepted as real data if pairwise correlation between the two series is within 0.015 of the desired value of $\rho$. The null pairs often had correlation generated spuriously through the simulation process. This was particularly problematic for generating small sample data, but setting, to a maximum of 0.5, for both true and null data reduced spurious correlation to manageable levels.
For particularly large time series, the AR(1) model can create particularly 'jagged' lines that may be unrealistic for many actual applications. To counter this, additional sets of data were created with a Hodrick-Prescott filter to capture the trend of the model and remove the noise in the cycle component. This filter creates a trend by penalising both errors in smoothness and in fit. The trend of the series, $\tau_{t}$ is defined by the equation:

$$\min_{\tau}(\sum_{t=1}^{T}(y_{t}-\tau_{t})^2 + \lambda \sum_{t=2}^{T-1}[(\tau_{t+1}-\tau_{t})-(\tau_{t}-\tau_{t-1})]^2)$$

Setting $\lambda = 1$ produced a smoothing that is more consistent with real world time series applications and did not introduce excessive spurious null correlation. However, for $t = 12, 24$, any form of smoothing was unfeasible without introducing extreme null correlation, hence smoothing was only used for $t = 48, 96$. Each generated lineup was produced as both a scatter plot and an overlaid line graph.

@@

\begin{itemize}
  \item Factor 1: Plot design, levels = scatter plot, overlaid time series
  \item Factor 2: Sample size (t), levels = 12, 24, 48, 96
  \item Factor 3: True plot correlation, $\rho$, levels = $\pm \{0.3, 0.5, 0.7, 0.9\}$
\end{itemize}

Additionally, there was a fourth factor for $t = 48, 96$:

@@

Each combination of factors was replicated in three different lineups, giving a total of 288 lineups (2 plots x 8 correlations x 6 sample sizes \& smoothness combinations). 
The position of the actual data plot in the lineup was randomized. The order that the factors are presented to a subject were also randomized, and each subject saw 5 scatter plots and 5 overlaid lines at a range of all absolute correlation levels. Each subject did not see the same data more than once. Subjects were asked to pick the plot in the lineup that has the strongest association, with basic examples of negatively and positively correlated plots. They also answered with reasons for choosing the plot, and the confidence that they have that this really is the plot showing the strongest correlation. Each subject evaluated 10 lineups, plus a further two trials to eliminate people attempted to game the system. If a trial was not answered correctly, the subject could not evaluate the remaining lineups. If it was answered correctly, the trial data was discarded and the subject moved onto the rest of the lineups. Earlier work by \citet{Majumder:2014} has found that repeated evaluations from a subject does not increase their ability to detect the true plot from a lineup, justifying our treatment of the results as independent evaluations. We recruited subjects from Amazon's Mechanical Turk service and ensured that each lineup was viewed by multiple subjects. In total there were 2088 lineup evaluations, however to better control each subject’s individual visual ability we removed data from subjects with less than ten individual evaluations leaving us with 1684 lineup evaluations for the analysis.

@@

\begin{figure}[htbp]
 \centering
<<results, dependson='setup', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, warning=FALSE, error=FALSE, message=FALSE>>=
Turk <- mutate(Turk, plot = ifelse(test_param =="line", "Overlaid Lines", "Scatterplot"), 
               sign=ifelse(correlation > 0, 1, 0), smoothed=ifelse(smoothed=="yes", 1, 0))
#ggplot(data=subset(Turk, smoothed==0), aes(x=correlation, y=percent, colour=plot)) + geom_smooth() + geom_point() + facet_grid(n~.) + scale_colour_discrete(name = "Plot Design") +  labs(title="Unsmoothed Plot Accuracy", y="Percent Correct", x="True Plot Correlation")
ggplot(data=Turk, aes(x=correlation, y=percent, colour=plot)) + geom_smooth(se=F) + geom_point() + facet_grid(n~smoothed) + 
  scale_colour_discrete(name = "Plot Design") + labs(title="Observed Data", y="Percent Correct", x="Correlation") +
  theme(legend.position="bottom")
@
  \label{results}
  \caption{Response accuracy for unsmoothed data, broken by unsmoothed (left), smoothed (right) and sample size. The dots are data points, the lines represent a naive model average fit. Both plot designs have similar performance for positively correlated data, however the overlaid line graphs are particularly weak at revealing negative correlation.}
\end{figure}

Figure~\ref{results} shows the raw results of the survey and Figure~\ref{powerdiff} shows the difference in power with a $95\%$ confidence interval calculated according to the formula in \ref{eq:2}. Values above zero indicate that the scatter plot had more power to detect the true plot than the overlaid lines.
Performance between the two plot designs appears to be similar for positive correlation, with the scatter plot power gradually improving relative to the overlaid lines as sample size increases. The scatter plot appears to be the superior choice as it retains its power with negative correlation, where the overlaid lines performance suffered. This is not surprising, as positive and negative correlation appear as symmetric graphs in the scatter plot but the two characteristics appear very differently in the overlaid line graph (Figure~\ref{designs}). The right side shows the performance benefits of smoothing overlaid line time series data. Much of the 'noise' in the time series is removed and studying the overall trend shows a greatly improved ability to successfully detect the true data plot when it is negatively correlated. The scatter plot still appears to be the superior plot, but much of the gap between the two is reduced for positive correlation.

\begin{figure}[htbp]
  \centering
<<powerdiff, dependson='results', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
Turk$correct <- Turk$percent*Turk$attempt
Turk$r <- round(Turk$correlation, 1)
#Turk.line <- subset(Turk, test_param=="line")
#Turk.scatter <- subset(Turk, test_param=="scatter")
prop <- Turk[order(Turk$r, Turk$n, Turk$smoothed),c("smoothed", "n", "attempt", "correct", "r", "plot")]
prop <- summarise(group_by(prop, smoothed, n, r), l.attempt=sum(ifelse(plot=="Overlaid Lines", attempt, 0)),
                                                  l.correct=sum(ifelse(plot=="Overlaid Lines", correct, 0)),
                                                  s.attempt=sum(ifelse(plot=="Scatterplot", attempt, 0)),
                                                  s.correct=sum(ifelse(plot=="Scatterplot", correct, 0)))
#Turk.line <- Turk.line[order(Turk.line$r, Turk.line$n, Turk.line$smoothed),c("smoothed", "n", "attempt", "correct", "r")]
#Turk.line <- summarise(group_by(Turk.line, smoothed, n, r, test_param), sum(attempt), sum(correct))
#Turk.scatter <- Turk.scatter[order(Turk.scatter$r, Turk.scatter$n, Turk.scatter$smoothed),c("smoothed", "n", "attempt", "correct", "r")]
#Turk.scatter <- summarise(group_by(Turk.scatter, smoothed, n, r), sum(attempt), sum(correct))
#colnames(Turk.scatter) <- c("smoothed", "n", "r", "s.attempt", "s.correct")
#colnames(Turk.line) <- c("smoothed", "n", "r", "l.attempt", "l.correct")
#prop <- cbind(Turk.line, Turk.scatter[4:5])
prop <- mutate(prop, l.adj = (l.correct+1)/(l.attempt+1), s.adj = (s.correct+1)/(s.attempt+1),
               df = s.attempt + l.attempt - 1, difference=s.adj-l.adj, t = qt(0.975, df),
               upper = difference + t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt),
               lower = difference - t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt))
ggplot(data=prop) + geom_errorbar(aes(x=r, ymax=upper, ymin=lower)) + geom_point(aes(x=r, y=difference)) +
        geom_hline(aes(yint=0, colour="red")) + labs(title="Difference in Power", x="Correlation", y="Scatterplot Advantage") + facet_grid(n~smoothed)
@
 \label{powerdiff}
 \caption{The difference in estimated power (scatter plot power minus overlaid line power) to reveal the true plot by smoothness and sample size with a 95\% confidence interval. The red line indicates the point where both plot designs are equally powerful while values above this line indicate that the scatter plot is more powerful than the overlaid line graph. Again, the largest advantage difference appears in negative correlation, where the scatter plot is more powerful.}
\end{figure}

\begin{figure}[htbp]
  \centering
<<samplesize, dependson='powerdiff', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
ggplot(data=subset(Turk, smoothed==0), aes(x=correlation, y=percent, colour=factor(n))) + geom_smooth(se=F) + geom_point() + facet_grid(plot~.) + scale_colour_discrete(name = "Sample size") +  labs(title="Effect of sample size, with unsmoothed data", y="Percent Correct", x="Correlation") + theme(legend.position="bottom")
@
 \label{samplesize}
 \caption{Replotting figure~\ref{results} to highlight the effects of changes in sample size of unsmoothed data. The scatter plot appears to universaly benefit from increased sample sizes but the overlaid lines only see a benefit in positive correlation.}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
<<time, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
full <- mutate(full, sign = ifelse(correlation >= 0, 1, 0), n24 = ifelse(n ==24, 1, 0), n48 = ifelse(n ==48,  1, 0), n96 = ifelse(n ==96, 1, 0), plot = ifelse(test_param ==1, "Overlaid Lines", "Scatterplot"))
time <- data.frame(plot=full$plot, time=full$log_time, correct=full$correct, rsgn=full$sign)
#ggplot(data=time, aes(x=time, fill=factor(correct))) +geom_bar(stat="bin") + facet_grid(~plot) +labs(x="Log(Time) Taken", y=element_blank()) + scale_fill_discrete(name = "Correct")
time$rsgn <- factor(time$rsgn, labels=c("r -","r +"))
tm <- summarise(group_by(time, correct, plot, rsgn), m=mean(time))
t1 <- ggplot(data=subset(time, correct==0), aes(x=time)) + geom_histogram(binwidth=0.5) + 
  geom_vline(data=subset(tm, correct==0), aes(xintercept = m), color="red") +
  facet_grid(plot~rsgn) + labs(title="Did not detect", x="Log(Time) Taken", y=element_blank())  
t2 <- ggplot(data=subset(time, correct==1), aes(x=time)) + geom_histogram(binwidth=0.5) + 
  geom_vline(data=subset(tm, correct==1), aes(xintercept = m), color="red") +
  facet_grid(plot~rsgn) + labs(title="Detected", x="Log(Time) Taken", y=element_blank())  
grid.arrange(t1, t2, ncol=2, top = "Log(Time) by plot design, true plot correlation and detection")
@
  \label{time} 
  \caption{Histograms of log time taken to make a decision. r+ and r- indicate the sign of true plot correlation. The vertical red line is the average time of the particular group. On average scatter plots are faster and the true plot is detected more often, with r-.}
\end{figure}

>>>>

<<setup, echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
library(lme4)
library(ggplot2)
library(mnormt)
library(tidyr)
library(knitr)
# Convert reshape2 to tidyr, melt to gather - done
#library(reshape2)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)
library(stringr)
#library(ipapi)
library(maps)
library(moments)
#library(pander)
#library(splitstackshape) - changed cSplit to unnest
#library(readr)
#library(lubridate)
# Convert these to read_csv at some point
correl <- read.csv("correlations.csv", stringsAsFactors = FALSE)
full <- read.csv("full.csv", row.names=1, stringsAsFactors = FALSE) 
Turk <- read.csv("turk.csv", stringsAsFactors = FALSE)

@@

To decide on an appropriate display requires awareness of cognitive principles in psychology such as the Gestalt law of common fate described by \citet{Wertheimer}. This law describes a tendency in human perception to 'see' a positive relationship between two objects that briefly move together over time, leading people to believe that they stay linked and share a 'common fate' (Figure~\ref{cny}), even if no relationship exists. \citet{Lee} work finds that humans are able to perceive a relationship between objects that move with temporal synchrony in accordance with the Gestalt Law, implying that a graphic than can utilise a dimension for time may be able to use the information provided by temporal dependence and potentially outperform the graph types that do not, such as the scatter plot. However, keep in mind magical thinking \citep{Diaconis},   the desire to find any pattern may lead to a false discovery of positive correlation. The law then has two effects, if positive correlation exists in temporal data, the observer may be drawn to it easier. If positive correlation does not exist, the observer may be fooled into thinking it does anyway. Either effect would lead to a more confident rejection of the null than if the same data was displayed as a scatter plot. Figure~\ref{cny} highlights the dangerous interaction between 'magical thinking' and the Gestalt Law, the observer is drawn to the period from April 8 to April 22, and may falsely believe that the two currencies may continue to move together. However, over the entire series the linear correlation between the two currencies is -0.12, indicating there is no actual relationship. Incorrectly relying on two currencies, or many other sets of economic variables, being dependent on each other can be extremely dangerous for a firm. \citet{Robbins} is similarly critical of the time-based line graph, where it is argued that information from overlaid line graphs can easily be misleading (Figure~\ref{murder}). When both lines dive after 1993, it is natural to compare the horizontal distance and say that the difference between the lines in 1994 is small. However, this is incorrect, as looking at the small horizontal distance is comparing the lines at different dates. It is not obvious that the difference in 1994 is the second greatest difference across the entire range, only slightly smaller than in 1993. \citet{Vanderplas} examines this effect in more detail. It is dubbed the 'Sine Illusion', where the human eye is poor at interpreting lines with such a sharp slope. As these problems are unique to the line graph, they argue that the scatter plot, which does not have any ability to display time, is the superior choice. However, in practice many data analysts are split between the two major alternatives for temporal data. The overlaid line graph is justified by its ability to present more information to the analyst; but many argue that it is this extra information that is misleading, and revert to the scatter plot for its strong non-temporal performance.

@@

\begin{equation}
\label{pvalue}
  \centering
  p-value = P(X \geq x | H_{0}) = 1 - B_{n, 1/m}(x-1)
\end{equation}

If all $K$ observers pick the same plot out of a lineup of twenty, we result in a p-value as small as $1/m^K$ . This research used $m= 20$, giving us a significance level (and Type 1 error rate) of $\alpha = 0.05$. The plot of 'true' data is placed randomly amongst 19 null plots to form the lineup.

If a subject selects multiple plots from the lineup, the p-value has a Bernoulli distribution with a probability $p_{j} = s/m$, where that observer made $s$ selections out of a lineup with $m$ plots \citep{Follett}. In the event that a lineup has observers making different numbers of selections, we have a binomial distribution with different success probabilities, and a Poisson Binomial distribution must be used to calculate the p-value. The probability mass function of a Poisson Binomial distribution can be written as:

\begin{equation} 
\label{pvalue2}
\centering
P(X=x) = \sum_{A \in T_{x}} \prod_{j \in A} p_{j} \prod_{j \in A^{c}} (1-p_{j})
\end{equation}

@@

As the power of a statistical test is defined as the probability of rejecting $H_{0}$ when it is false, the power of a lineup test is viewed as the probability of detecting the true plot. We approximate the power as $\hat\pi = x/K$, where $x$ observers out of $K$ correctly detected the true plot out of the lineup.
We can them estimate the power difference of competing lineups, $\hat\pi_{1} - \hat\pi_{2}$. An $\alpha \times 100\%$ confidence interval is calculated as \citep{Hofmann:2012}:

\begin{equation} \label{propci}
  \centering
  \hat\pi_{1} - \hat\pi_{2}\pm t_{1-\alpha,2n-1}\sqrt{\hat\pi_{1} (1-\hat\pi_{1}) /n_{1} + \hat\pi_{2} (1-\hat\pi_{2}) /n_{2} }
\end{equation}

@@

Some individuals may have a better natural ability at detecting the correct correlation pattern in the graphics, but by having each participant viewing multiple lineups and each lineup being viewed by multiple different participants, this can be controlled via a random effect variable in the model. We recruited participants through Amazon's Mechanical Turk, where the lineup protocol has been applied to a number of problems in prior papers \citep{Heer, Majumder:2013, Wickham}. Amazon’s Mechanical Turk \citep{MTurk} is a labour crowd-sourcing platform developed to give easy access to workers with basic tasks paid in line with the United States minimum wage. MTurk can be used to recruit subjects to read a variety of graphics and report on particular visual tasks. The time taken to decide, confidence in the decision and the reason for that decision can also be recorded. \citet{Heer} use MTurk to replicate previous findings in graphical perception from \citet{Cleveland} and find that it is a valid method of data collection, once controls to eliminate anyone 'gaming' the system by randomly selecting answers to minimise time spent working are implemented. Further details of our use of Amazons’ Mechanical Turk is included in Section 3. The protocol has successfully been used to measure statistical power as a means to determine plot type superiority in \citet{Hofmann:2012}, and this work follows that approach.

@@

$$~\mathcal{N}\left\(\mu= \left| \begin{array}{c}0 \\0 \end{array} \right|,\hspace{4mm}\Sigma = \left| \begin{array}{cc}1 & \rho \\ \rho & 1\end{array} \right|\right\)$$

The variation in the covariance parameter  allows for control in the correlation in the pairs of data. The null data has no structure, so uncorrelated pairs are produced by setting $\rho = 0$. True data pairs are generated with $\rho = \pm \{0.3, 0.5, 0.7, 0.9\}$. Pilot testing found that the rate of picking the true data plot fell to approximately $5\%$, the $1/m$ rate of randomly picking the true plot, for pairs with correlation below $\rho = 0.3$. Further to this, relationships weaker than this point are often of little interest to analysts. Hence we found it unnecessary to test a more full range of correlations. Simulating data does not produce correlations exactly equal to $\rho$, so imulated data pairs were accepted as real data if pairwise correlation between the two series is within 0.015 of the desired value of $\rho$. The null pairs often had correlation generated spuriously through the simulation process. This was particularly problematic for generating small sample data, but setting $\beta = max(0.5,100t)$ for both true and null data reduced spurious correlation to manageable levels.
For particularly large time series, the AR(1) model can create particularly 'jagged' lines that may be unrealistic for many actual applications. To counter this, additional sets of data were created with a Hodrick-Prescott filter to capture the trend of the model and remove the noise in the cycle component. This filter creates a trend by penalising both errors in smoothness and in fit. The trend of the series, $\tau_{t}$ is defined by the equation:

\begin{equation}
\centering
\label{hpfilter}
\min_{\tau}(\sum_{t=1}^{T}(y_{t}-\tau_{t})^2 + \lambda \sum_{t=2}^{T-1}[(\tau_{t+1}-\tau_{t})-(\tau_{t}-\tau_{t-1})]^2)
\end{equation}

Setting $\lambda = 1$ produced a smoothing that is more consistent with real world time series applications and did not introduce excessive spurious null correlation. However, for $t = \{12, 24\}$, any form of smoothing was unfeasible without introducing extreme null correlation, hence smoothing was only used for $t = \{48, 96\}$. Each generated lineup was produced as both a scatter plot and an overlaid line graph.

@@

\begin{itemize}
  \item Factor 1: Plot design, levels = scatter plot, overlaid time series
  \item Factor 2: Sample size (t), levels = \{12, 24, 48, 96\}
  \item Factor 3: True plot correlation, $\rho$, levels = $\pm \{0.3, 0.5, 0.7, 0.9\}$
\end{itemize}

Additionally, there was a fourth factor for $t = \{48, 96\}$:

@@

Each combination of factors was replicated in three different lineups, giving a total of 288 lineups (3 replications x 2 plots x 8 correlations x 6 sample size and smoothness combinations). 
The position of the actual data plot in the lineup was randomized. The order that the factors are presented to a subject were also randomized, and each subject saw 5 scatter plots and 5 overlaid lines at a range of all absolute correlation levels. Each subject did not see the same data more than once. Subjects were asked to pick the plot in the lineup that has the strongest association, with basic examples of negatively and positively correlated plots. They also answered with reasons for choosing the plot, and the confidence that they have that this really is the plot showing the strongest correlation. Each subject evaluated 10 lineups, plus a further two trials to eliminate people attempted to game the system. If a trial was not answered correctly, the subject could not evaluate the remaining lineups. If it was answered correctly, the trial data was discarded and the subject moved onto the rest of the lineups. Earlier work by \citet{Majumder:2014} has found that repeated evaluations from a subject does not increase their ability to detect the true plot from a lineup, justifying our treatment of the results as independent evaluations. We recruited subjects from Amazon's Mechanical Turk service and ensured that each lineup was viewed by multiple subjects. In total there were 2088 lineup evaluations, however to better control each subject’s individual visual ability we removed data from subjects with less than ten individual evaluations leaving us with 1684 lineup evaluations for the analysis.

@@

\begin{figure}[htbp]
 \centering
<<results, dependson='setup', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, warning=FALSE, error=FALSE, message=FALSE>>=
Turk <- mutate(Turk, plot = ifelse(test_param =="line", "Overlaid Lines", "Scatterplot"), 
               sign=ifelse(correlation > 0, 1, 0), smoothed=ifelse(smoothed=="yes", 1, 0))
#ggplot(data=subset(Turk, smoothed==0), aes(x=correlation, y=percent, colour=plot)) + geom_smooth() + geom_point() + facet_grid(n~.) + scale_colour_discrete(name = "Plot Design") +  labs(title="Unsmoothed Plot Accuracy", y="Percent Correct", x="True Plot Correlation")
ggplot(data=Turk, aes(x=correlation, y=percent, colour=plot)) + geom_smooth(se=F) + geom_point() + facet_grid(n~smoothed) + 
  scale_colour_discrete(name = "Plot Design") + labs(y="True plot detection rate", x="Correlation") +
  theme(legend.position="bottom")
@
  \label{results}
  \caption{Response accuracy for unsmoothed data, broken by unsmoothed (left), smoothed (right) and sample size. The dots are data points, the lines represent a naive model average fit. Both plot designs have similar performance for positively correlated data, however the overlaid line graphs are particularly weak at revealing negative correlation.}
\end{figure}

Figure~\ref{results} shows the raw results of the survey and Figure~\ref{powerdiff} shows the difference in power with a $95\%$ confidence interval calculated according to  formula~\ref{propci}. Values above zero indicate that the scatter plot had more power to detect the true plot than the overlaid lines.
Performance between the two plot designs appears to be similar for positive correlation, with the scatter plot power gradually improving relative to the overlaid lines as sample size increases. The scatter plot appears to be the superior choice as it retains its power with negative correlation, where the overlaid lines performance suffered. This is not surprising, as positive and negative correlation appear as symmetric graphs in the scatter plot but the two characteristics appear very differently in the overlaid line graph (Figure~\ref{designs}). The right side shows the performance benefits of smoothing overlaid line time series data. Much of the 'noise' in the time series is removed and studying the overall trend shows a greatly improved ability to successfully detect the true data plot when it is negatively correlated. The scatter plot still appears to be the superior plot, but much of the gap between the two is reduced for positive correlation.

\begin{figure}[htbp]
  \centering
<<powerdiff, dependson='results', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
Turk$correct <- Turk$percent*Turk$attempt
Turk$r <- round(Turk$correlation, 1)
#Turk.line <- subset(Turk, test_param=="line")
#Turk.scatter <- subset(Turk, test_param=="scatter")
prop <- Turk[order(Turk$r, Turk$n, Turk$smoothed),c("smoothed", "n", "attempt", "correct", "r", "plot")]
prop <- summarise(group_by(prop, smoothed, n, r), l.attempt=sum(ifelse(plot=="Overlaid Lines", attempt, 0)),
                                                  l.correct=sum(ifelse(plot=="Overlaid Lines", correct, 0)),
                                                  s.attempt=sum(ifelse(plot=="Scatterplot", attempt, 0)),
                                                  s.correct=sum(ifelse(plot=="Scatterplot", correct, 0)))
#Turk.line <- Turk.line[order(Turk.line$r, Turk.line$n, Turk.line$smoothed),c("smoothed", "n", "attempt", "correct", "r")]
#Turk.line <- summarise(group_by(Turk.line, smoothed, n, r, test_param), sum(attempt), sum(correct))
#Turk.scatter <- Turk.scatter[order(Turk.scatter$r, Turk.scatter$n, Turk.scatter$smoothed),c("smoothed", "n", "attempt", "correct", "r")]
#Turk.scatter <- summarise(group_by(Turk.scatter, smoothed, n, r), sum(attempt), sum(correct))
#colnames(Turk.scatter) <- c("smoothed", "n", "r", "s.attempt", "s.correct")
#colnames(Turk.line) <- c("smoothed", "n", "r", "l.attempt", "l.correct")
#prop <- cbind(Turk.line, Turk.scatter[4:5])
prop <- mutate(prop, l.adj = (l.correct+1)/(l.attempt+1), s.adj = (s.correct+1)/(s.attempt+1),
               df = s.attempt + l.attempt - 1, difference=s.adj-l.adj, t = qt(0.975, df),
               upper = difference + t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt),
               lower = difference - t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt))
ggplot(data=prop) + geom_errorbar(aes(x=r, ymax=upper, ymin=lower)) + geom_point(aes(x=r, y=difference)) +
        geom_hline(aes(yint=0, colour="red")) + labs(x="Correlation", y="Scatterplot power minus Overlaid Lines") + facet_grid(n~smoothed)
@
 \label{powerdiff}
 \caption{The difference in estimated power (scatter plot power minus overlaid line power) to reveal the true plot by smoothness and sample size with a 95\% confidence interval. The red line indicates the point where both plot designs are equally powerful while values above this line indicate that the scatter plot is more powerful than the overlaid line graph. The largest differences appear in negative correlation, where the scatter plot is more powerful.}
\end{figure}

\begin{figure}[htbp]
  \centering
<<samplesize, dependson='powerdiff', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
ggplot(data=subset(Turk, smoothed==0), aes(x=correlation, y=percent, colour=factor(n))) + geom_smooth(se=F) + geom_point() + facet_grid(plot~.) + scale_colour_discrete(name = "Sample size") +  labs(y="Percent Correct", x="Correlation") + theme(legend.position="bottom")
@
 \label{samplesize}
 \caption{Replotting figure~\ref{results} to highlight the effects of changes in sample size of unsmoothed data. The scatter plot appears to universaly benefit from increased sample sizes but the overlaid lines only see a benefit in positive correlation.}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
<<time, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
full <- mutate(full, sign = ifelse(correlation >= 0, 1, 0), n24 = ifelse(n ==24, 1, 0), n48 = ifelse(n ==48,  1, 0), n96 = ifelse(n ==96, 1, 0), plot = ifelse(test_param ==1, "Overlaid Lines", "Scatterplot"))
time <- data.frame(plot=full$plot, time=full$log_time, correct=full$correct, rsgn=full$sign)
#ggplot(data=time, aes(x=time, fill=factor(correct))) +geom_bar(stat="bin") + facet_grid(~plot) +labs(x="Log(Time) Taken", y=element_blank()) + scale_fill_discrete(name = "Correct")
time$rsgn <- factor(time$rsgn, labels=c("r -","r +"))
tm <- summarise(group_by(time, correct, plot, rsgn), m=mean(time))
t1 <- ggplot(data=subset(time, correct==0), aes(x=time)) + geom_histogram(binwidth=0.5) + 
  geom_vline(data=subset(tm, correct==0), aes(xintercept = m), color="red") +
  facet_grid(plot~rsgn) + labs(title="Did not detect", x="Log(Time) Taken", y=element_blank())  
t2 <- ggplot(data=subset(time, correct==1), aes(x=time)) + geom_histogram(binwidth=0.5) + 
  geom_vline(data=subset(tm, correct==1), aes(xintercept = m), color="red") +
  facet_grid(plot~rsgn) + labs(title="Detected", x="Log(Time) Taken", y=element_blank())  
grid.arrange(t1, t2, ncol=2)
@
  \label{time} 
  \caption{Histograms of log time taken to make a decision. r+ and r- indicate the sign of true plot correlation. The vertical red line is the average time of the particular group. On average scatter plots are faster and the true plot is detected more often, with r-.}
\end{figure}
